\TITLE

### 2020-12-18 vonLaszewski-covid.tex
update with the new template

\documentclass[12pt]{article}

\usepackage{amsfonts,amsmath,amssymb,amsthm}
\usepackage{array}
\usepackage{authblk}
\usepackage{color}
\usepackage[table]{xcolor}
\usepackage{epsfig,fancyhdr}
\usepackage{geometry,graphicx}
\usepackage[]{hyperref} 
\usepackage{indentfirst}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{listings}
\usepackage{natbib}
\usepackage{rotating}
\usepackage{comment}
\usepackage{booktabs}
\usepackage{todonotes}
\setuptodonotes{inline}
\usepackage{makecell}
\usepackage{enumitem}
\usepackage{tcolorbox}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{sectsty,secdot}
\sectionfont{\bf\fontsize{14}{2em}\selectfont}
\subsectionfont{\bf\fontsize{12}{2em}\selectfont}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\geometry{left=2.5cm,right=2cm,top=2.5cm,bottom=1.5cm}
\setlength{\parindent}{0.63cm}

\setcounter{page}{1}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{remark}{Remark}
\pagestyle{empty}
\usepackage{lineno}
\setcitestyle{square}

\newcommand*\rot{\rotatebox{90}}

%---------page header---------%

%\fancyhead[RO]{odd number page}
%\fancyhead[LE]{even number page}

\pagestyle{plain} 
\linenumbers

%-----------------------------%
\newtheorem{question}{Question}

\newtheorem{requirement}{Requirement}

@@

\newtcolorbox{redbox}{colback=red!5!white,colframe=red!75!black,width=3cm}
\newtcolorbox{yellowbox}{colback=yellow!5!white,colframe=yelloe!75!black,width=3cm}

\newcommand{\DONE}{{\color{green!60!black}\makebox[0pt][l]{$\square$}\raisebox{.15ex}{\hspace{0.1em}$\checkmark$}}~}
\newcommand{\DOIT}{{\color{red!60!black}\makebox[0pt][l]{$\square$}\raisebox{.15ex}{\hspace{0.1em}$\boxtimes$}}~}
\newcommand{\PROGRESS}{(in progress)~}

@@

\input{response-v3-content}

\title{{\bf\fontsize{14pt}{18pt}\selectfont AICov: An Integrative Deep Learning Framework for COVID-19 Forecasting with Population Covariates}}
\author[1]{{\fontsize{12pt}{0.5em}\selectfont Geoffrey C. Fox}}
\author[1]{{\fontsize{12pt}{18pt}\selectfont Gregor von Laszewski}}
\author[1]{{\fontsize{12pt}{0.5em}\selectfont Fugang Wang}}
\author[2,3]{{\fontsize{12pt}{0.5em}\selectfont Saumyadipta Pyne}}

\affil[1]{{\fontsize{12pt}{0.5em}\selectfont Digital Science Center, Indiana University, Bloomington Indiana, USA}}
\affil[2]{{\fontsize{12pt}{0.5em}\selectfont Public Health Dynamics Lab, and Department of Biostatistics, University of Pittsburgh, Pittsburgh, Pennsylvania, USA}}
\affil[3]{{\fontsize{12pt}{0.5em}\selectfont Health Analytics Network, Pennsylvania, USA}}

\newcommand{\ABSTRACT}{
The COVID-19 pandemic has had profound global consequences on health,
economic, social, behavioral, and almost every major aspect of human
life. Therefore, it is of great importance to model COVID-19 and other
pandemics in terms of the broader social contexts in which they take
place.  We present the architecture of an artificial intelligence
enhanced COVID-19 analysis (in short AICov) , which provides an
integrative deep learning framework for COVID-19 forecasting with
population covariates, some of which may serve as putative risk
factors.  We have integrated multiple different strategies into AICov,
including the ability to use deep learning strategies based on LSTM
and event modeling.  To demonstrate our approach, we have introduced a
framework that integrates population covariates from multiple
sources. Thus, AICov not only includes data on COVID-19 cases and
deaths but, more importantly, the population's socioeconomic, health,
and behavioral risk factors at a local level. The compiled data are
fed into AICov, and thus we obtain improved prediction by the
integration of the data to our model as compared to one that only uses
case and death data. As we use deep learning our models adapt over
time while learning the model from past data.
}
\date{}

\newcommand{\TODO}[1]{
\todo[color=yellow!20]{#1}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\baselinestretch}{1.07}
\renewcommand{\thefootnote}{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\maketitle
\thispagestyle{empty}

\begin{quotation}
\centerline{\bfseries{\fontsize{14pt}{1em}\selectfont Abstract}}\par
\setlength{\parindent}{3.22cm}
{\fontsize{12pt}{1.07em}\selectfont~~\ABSTRACT}

\vspace{12pt}
\noindent {\bfseries{Key words and phrases:}}
COVID-19, deep learning, prediction, risk factors, comorbidities.
\par
\end{quotation}\par

\def\thefigure{\arabic{figure}}
\def\thetable{\arabic{table}}

\renewcommand{\theequation}{\thesection.\arabic{equation}}

\fontsize{12}{1.07em}\selectfont

\setcounter{section}{0} %***
\setcounter{equation}{0} %-1

@@

The first known case was traced back to 17 November 2019. Since the
emergence of a cluster of cases in Wuhan, China, on 31 December 2019,
COVID-19 has spread rapidly worldwide. Just six months later, as of 23
June 2020, there are 8,993,659 COVID-19 cases and 469,587 deaths
globally. On 30 January 2020, the World Health Organization (WHO)
\cite{www-who} declared COVID-19 to be a Public Health Emergency of
International Concern and subsequently, on 11 March 2020, a pandemic.

In the United States, the first case of COVID-19 was confirmed on 20
January 2020 in Washington State, and as of 23 June 2020, there were
over 2.3 million confirmed cases and 121,167 deaths
\cite{www-nyt-map}.  Although initial efforts to reduce the spread
took place, recent data shows a resurgence at an increased scale. In
the absence of a vaccine or treatment to effectively combat the
disease, non-pharmaceutical policy interventions such as social
distancing and lockdowns are recommended by health experts to prevent
further transmission. To gain insights into the possible impact of
such measures on COVID-19 outcomes, we depend upon the ability to
accurately forecast the spread of reported cases and confirmed deaths
and recoveries. Naturally, the accuracy of forecasting relies on the
availability of current, reliable data and historical data to
determine estimates of uncertainty.

During outbreaks of epidemics, initially data on cases and deaths
could be scarce, and the quality of data annotation, validation, and
aggregation might be uncertain. For instance, changes in clinical data
entry, such as the addition of a new category clinically diagnosed to
the existing lab-confirmed category, may likely have been reflected in
the reporting \cite{Petropoulos2020-hh}. Initial fears about COVID-19
among the general population with regards to the trajectory of the
pandemic could also affect administrative reporting. Diverse media
sources and differences in local and federal policies may add to the
general uncertainty about disease progression. Nevertheless, a
data-driven approach to forecasting can offer valuable insights into
the disease dynamics, and thereby an ability to objectively plan for
the near future.

Unlike earlier global outbreaks, during COVID-19, the current
worldwide digital ecosystem allows for real-time data collection,
which in turn is used by artificial intelligence (AI) and deep
learning systems to understand healthcare trends, model risk
associations, and predict outcomes \cite{Ting2020-lw}.  In addition to
traditional public health surveillance strategies available to most
countries, a variety of static and dynamic data types may be
integrated to model the scale and dynamics of this pandemic
\cite{Pyne2015-ao}. Several organizations, including the Johns Hopkins
University’s Center for Systems Science and Engineering, the New York
Times, the Atlantic’s COVID-19 tracking project, among others have
developed real-time tracking maps for following COVID-19 cases around
the world using data from the U.S.~Centers for Disease Control and
Prevention (CDC), WHO, and other international health
agencies. Notably, the CDC receives forecasts from several modeling
groups that use a wide variety of approaches ranging from SEIR
(Susceptible-Exposed-Infectious-Recovered) to Agent-based to Bayesian
models to understand the non-pharmaceutical policy interventions (or
the lack thereof) to predict disease dynamics and its impact on human
lives \cite{www-cdc-modeling-forecast}.

@@

\begin{comment}
\begin{verbatim}
    model.add(Dense(InitialMLP, activation=activationvalue, input_shape=(n_timesteps, n_features)))

    model.add(LSTM(n_nodes, unroll=True, recurrent_dropout= recurrent_dropout1, dropout = dropout1,
                   activation= activationvalue , return_sequences=True, recurrent_activation="sigmoid",
                   input_shape=(n_timesteps, nextround)))

    model.add(LSTM(n_nodes, unroll=True, recurrent_dropout= recurrent_dropout2, dropout = dropout2,
               activation= activationvalue, recurrent_activation="sigmoid",
               input_shape=(n_timesteps, n_nodes)))
\end{verbatim}

\begin{verbatim}
Window Size  1  
n_samples: 12540 
n_timesteps: 1 
n_features: 3 
n_outputs: 30 
batch_size: 110 
n_nodes: 16 
epochs: 200
is NaN  77000  
percent  20.47  
not NaN  299200
\end{verbatim}
\end{comment}

@@

%%%%%%

@@

\begin{comment}
\begin{figure}[!h]
    \centering
    \includegraphics[width=1.0\textwidth]{images/boxwhisker/swarmplot_cases.pdf}
    \caption{Caption}
    \label{fig:my_label}

    \centering
    \includegraphics[width=1.0\textwidth]{images/boxwhisker/swarmplot_death.pdf}
    \caption{Caption}
    \label{fig:my_label}
\end{figure}
\end{comment}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{comment}

@@

\vskip 14pt
\noindent {\bfseries{\fontsize{14pt}{1em}\selectfont Supplementary Materials}}

Contain
the brief description of the online supplementary materials.
\par

\end{comment}
%
% ACKNOWLEDGMENT
%

\section{Acronyms}

This section contains the list of acronyms used in the paper for easy
reference. The definitions of the risk factors are however given
within the paper in Table~\ref{tab:risk-factors} on page
\pageref{tab:risk-factors} and not repeated here.

\begin{itemize}[leftmargin=5cm]
\item[{\bf API:}] Application Programming Interfaces
\item[{\bf BRFSS:}] Behavioral Risk Factor Surveillance System
\item[{\bf CDC:}] Center for Desease Control
\item[{\bf CIF21:}] Cyberinfrastructure Framework for 21st Century 
\item[{\bf COVID-19:}] COrona VIrus Disease 2019 
\item[{\bf CSSE:}] Center for Systems Science and Engineering
\item[{\bf DIBBS:}] Data Infrastructure Building Blocks
\item[{\bf FIPS:}] Federal Information Processing Standards
\item[{\bf GPU:}] Graphics Processing Unit 
\item[{\bf HPC:}] High Performance Computing
\item[{\bf LSTM:}] Long Short-Term Memory
\item[{\bf NBDIF:}] NIST Big Data Working Group
\item[{\bf NCHS:}] CDC National Center for Health Statistics
\item[{\bf NIST:}] National Institute of Standards and Technology
\item[{\bf RELU:}] REctified Linear Unit
\item[{\bf REST:}] REpresentational State Transfer 
\item[{\bf RNN:}] Recurrent Neural Network
\item[{\bf SEIR:}] Susceptible, Expose, Infectious, Recovered)
\item[{\bf SIR}] Susceptible, Infectious, Recovered
\item[{\bf U.S.:}] United States of America
\item[{\bf NSF:}] National Science Foundation
\item[{\bf MSE:}] Mean Squared Error
\item[{\bf SELU:}] Scaled Exponential Linear Unit
\item[{\bf CI:}] Cyber Infrastructure
\item[{\bf CINES:}] CyberInfrastructure for Network Engineering and Science
\item[{\bf WHO:}] World Health Organization
\item[{\bf NONE:}] \todo{NO FACTORS}
\item[{\bf NORM\_POP:}] Normalized Population
\item[{\bf cum:}] cumulative
\end{itemize}

\vskip 14pt
\noindent {\bfseries{\fontsize{14pt}{1em}\selectfont Acknowledgements}}

@@

\par

%
% BIBLIOGRAPHY
%

%\iffalse
\fontsize{11}{1em}\selectfont
\renewcommand\bibname{\bfseries{\fontsize{14pt}{1em}\selectfont References}}
\expandafter\ifx\csname
natexlab\endcsname\relax\def\natexlab#1{#1}\fi
\expandafter\ifx\csname url\endcsname\relax
  \def\url#1{\texttt{#1}}\fi
\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL}\fi
%\fi

%\clearpage

%\bibliographystyle{plainurl}
%\bibliographystyle{plain}
\bibliographystyle{plainnat}

\Bibliography{bib/data-covid19,bib/lstm,bib/cloudmesh}

>>>>

\documentclass[letterpaper, inpress]{jds} % use this for production

%\documentclass[a4paper, review]{jds}      % use this for review
\usepackage{lineno}
\linenumbers

\usepackage{calc}
\usepackage{comment}

% color management

\usepackage{color}
\usepackage{tcolorbox}

% table managemet

\usepackage[table]{xcolor}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{rotating, multirow}

% Avoid non breaking words

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% the following edits should be done by Journal typesetters
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setcounter{page}{1}            % set the first page number
\jdsmonth{January}              % month
\jdsyear{2021}                  % year
\jdsvolume{xx}                  % volume number
\jdsissue{xx}                   % issue number
\jdsdoi{xx.xxxx/xxxxxxxxx}      % doi
\jdsreceived{???, 2020}       % optional: comment it out if no received date
\jdsaccepted{???, 2021}         % optional: comment it out if no accepted date
%% manually set running header for a shorter list of authors if needed
% \shortauthors{A Author A, et al.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% edits by authors are given below
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amsfonts,amsmath,amssymb,amsthm}
\usepackage{booktabs}

\newcommand{\TITLE}{AICov: An Integrative Deep
    Learning Framework for COVID-19 Forecasting with Population
    Covariates}

\newcommand{\ABSTRACT}{
The COVID-19 pandemic has had profound global consequences on health,
economic, social, behavioral, and almost every major aspect of human
life. Therefore, it is of great importance to model COVID-19 and other
pandemics in terms of the broader social contexts in which they take
place.  We present the architecture of an artificial intelligence
enhanced COVID-19 analysis (in short AICov) , which provides an
integrative deep learning framework for COVID-19 forecasting with
population covariates, some of which may serve as putative risk
factors.  We have integrated multiple different strategies into AICov,
including the ability to use deep learning strategies based on LSTM
and event modeling.  To demonstrate our approach, we have introduced a
framework that integrates population covariates from multiple
sources. Thus, AICov not only includes data on COVID-19 cases and
deaths but, more importantly, the population's socioeconomic, health,
and behavioral risk factors at a local level. The compiled data are
fed into AICov, and thus we obtain improved prediction by the
integration of the data to our model as compared to one that only uses
case and death data. As we use deep learning our models adapt over
time while learning the model from past data.
}

\title[AICov: Deep
    Learning Framework for COVID-19 Forecasting with Population
    Covariates]{\TITLE}

\author[1]{Geoffrey C. Fox}
\author[1]{Gregor von Laszewski\thanks{Corresponding author. Email: laszewski@gmail.com}}
\author[1]{Fugang Wang}
\author[2,3]{Saumyadipta Pyne}

\affil[1]{Digital Science Center, Indiana University, Bloomington Indiana, USA}
\affil[2]{Public Health Dynamics Lab, and Department of Biostatistics, University of Pittsburgh, Pittsburgh, Pennsylvania, USA}
\affil[3]{Health Analytics Network, Pennsylvania, USA}

@@

\maketitle

\begin{abstract}
  \ABSTRACT
\end{abstract}

\begin{keywords} % alphabetical; excluding anything in the title already
  Cloudmesh;
  Comorbidities;
  Prediction;
  Risk factors.
\end{keywords}

@@

The first known case was traced back to 17 November 2019. Since the
emergence of a cluster of cases in Wuhan, China, on 31 December 2019,
COVID-19 has spread rapidly worldwide. Just six months later, as of 23
June 2020, there are 8,993,659 COVID-19 cases and 469,587 deaths
globally. On 30 January 2020, the World Health Organization (WHO)
\citep{www-who} declared COVID-19 to be a Public Health Emergency of
International Concern and subsequently, on 11 March 2020, a pandemic.

In the United States, the first case of COVID-19 was confirmed on 20
January 2020 in Washington State, and as of 23 June 2020, there were
over 2.3 million confirmed cases and 121,167 deaths
\citep{www-nyt-map}.  Although initial efforts to reduce the spread
took place, recent data shows a resurgence at an increased scale. In
the absence of a vaccine or treatment to effectively combat the
disease, non-pharmaceutical policy interventions such as social
distancing and lockdowns are recommended by health experts to prevent
further transmission. To gain insights into the possible impact of
such measures on COVID-19 outcomes, we depend upon the ability to
accurately forecast the spread of reported cases and confirmed deaths
and recoveries. Naturally, the accuracy of forecasting relies on the
availability of current, reliable data and historical data to
determine estimates of uncertainty.

During outbreaks of epidemics, initially data on cases and deaths
could be scarce, and the quality of data annotation, validation, and
aggregation might be uncertain. For instance, changes in clinical data
entry, such as the addition of a new category clinically diagnosed to
the existing lab-confirmed category, may likely have been reflected in
the reporting \citep{Petropoulos2020-hh}. Initial fears about COVID-19
among the general population with regards to the trajectory of the
pandemic could also affect administrative reporting. Diverse media
sources and differences in local and federal policies may add to the
general uncertainty about disease progression. Nevertheless, a
data-driven approach to forecasting can offer valuable insights into
the disease dynamics, and thereby an ability to objectively plan for
the near future.

Unlike earlier global outbreaks, during COVID-19, the current
worldwide digital ecosystem allows for real-time data collection,
which in turn is used by artificial intelligence (AI) and deep
learning systems to understand healthcare trends, model risk
associations, and predict outcomes \citep{Ting2020-lw}.  In addition to
traditional public health surveillance strategies available to most
countries, a variety of static and dynamic data types may be
integrated to model the scale and dynamics of this pandemic
\citep{Pyne2015-ao}. Several organizations, including the Johns Hopkins
University’s Center for Systems Science and Engineering, the New York
Times, the Atlantic’s COVID-19 tracking project, among others have
developed real-time tracking maps for following COVID-19 cases around
the world using data from the U.S.~Centers for Disease Control and
Prevention (CDC), WHO, and other international health
agencies. Notably, the CDC receives forecasts from several modeling
groups that use a wide variety of approaches ranging from SEIR
(Susceptible-Exposed-Infectious-Recovered) to Agent-based to Bayesian
models to understand the non-pharmaceutical policy interventions (or
the lack thereof) to predict disease dynamics and its impact on human
lives \citep{www-cdc-modeling-forecast}.

@@

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\section*{Supplementary Materials}

Contain
the brief description of the online supplementary materials.

%
% ACKNOWLEDGMENT
%

\section*{Acknowledgements}

@@

\clearpage

\section*{Acronyms}

This section contains the list of acronyms in Table{tab:acronymes} as used in the paper for easy
reference. The definitions of the risk factors are however given
within the paper in Table~\ref{tab:risk-factors} on page
\pageref{tab:risk-factors} and not repeated here.

\begin{table}[!h]
  \begin{tabular}{ll}
    \toprule
    Abbreviation & Description \\
    \midrule
    API & Application Programming Interfaces \\
    BRFSS & Behavioral Risk Factor Surveillance System \\
    CDC & Center for Desease Control \\
    CI & Cyber Infrastructure \\
    CIF21 & Cyberinfrastructure Framework for 21st Century  \\
    CINES & CyberInfrastructure for Network Engineering and Science \\
    COVID-19 & COrona VIrus Disease 2019  \\
    CSSE & Center for Systems Science and Engineering \\
    cum & cumulative \\
    DIBBS & Data Infrastructure Building Blocks \\
    FIPS & Federal Information Processing Standards \\
    GPU & Graphics Processing Unit  \\
    HPC & High Performance Computing \\
    LSTM & Long Short-Term Memory \\
    MSE & Mean Squared Error \\
    NBDIF & NIST Big Data Working Group \\
    NCHS & CDC National Center for Health Statistics \\
    NIST & National Institute of Standards and Technology \\
    NONE & \todo{NO FACTORS} \\
    NORM\_POP & Normalized Population \\
    NSF & National Science Foundation \\
    RELU & REctified Linear Unit \\
    REST & REpresentational State Transfer  \\
    RNN & Recurrent Neural Network \\
    SEIR & Susceptible, Expose, Infectious, Recovered) \\
    SELU & Scaled Exponential Linear Unit \\
    SIR &  Susceptible, Infectious, Recovered \\
    U.S. & United States of America \\
    WHO & World Health Organization \\
    \midrule
\end{tabular}
\end{table}

\clearpage

\bibliographystyle{jds}
\bibliography{bib/data-covid19,bib/lstm,bib/cloudmesh}
